---
title: Object oriented bipartite
author: 
  - name: Tobias Bauer
    affiliation: Animal Network Ecology, University of Hamburg
    email: tobias.bauer-2@uni-hamburg.de
date: "`r format(Sys.time(), '%B %d, %Y')`"
abstract: This vignette provides an overview of the new object oriented approach of the bipartite package.
          It includes multiple examples illustrating the intention behind various function arguments.
toc-title: "Contents"
output: 
  rmarkdown::html_vignette:
    toc: true  
    fig_width: 7
    fig_height: 7
editor_options: 
  chunk_output_type: inline
vignette: >
  %\VignetteIndexEntry{Object oriented bipartite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Creating S3-objects in bipartite

Starting with version XX.XX the `bipartite`-package enables object-oriented programming (OOP) using the S3 system. To understand what that mean we will demonstrate the new functionalities with the aid of the dataset `Safariland`.

```{r, results = "hide", message = FALSE, cache = FALSE}
library(bipartite)
data(Safariland)
options(max.print = 10)
```



```{r}
class(Safariland)
str(Safariland)
```

The dataset is an integer matrix of size $9\times 27$ with named rows and columns. From the help file we know that the columns represent pollinators and the rows represent plant species. 

Sometimes, that's all the information available about a network. More often, however, additional information is available about individual nodes or the network as a whole. 

```{r}
plant_attributes <- read.csv("plant_attributes.csv", row.names = 1)
plant_attributes <- plant_attributes[rownames(Safariland), ]
plant_attributes
```

```{r}
pollinator_attributes <- read.csv("pollinator_attributes.csv", row.names = 1)
pollinator_attributes <- pollinator_attributes[colnames(Safariland), ]
pollinator_attributes
```

```{r}
Safariland.web <- bipartite_web(Safariland,
                                lower_attributes = plant_attributes,
                                higher_attributes = pollinator_attributes,
                                meta_attributes = list("type" = "pollination"))
class(Safariland.web)
```

```{r}
print(Safariland.web, max = 10)
plot(Safariland.web)
```

```{r}
attributes(Safariland.web)
```

# Accessing the attributes
```{r}
meta_attributes(Safariland.web)$start_year <- 1999
meta_attributes(Safariland.web)$end_year <- 2000
# higher_attributes(Safariland.web) <- data.frame(rand = runif(ncol(Safariland.web)), row.names = colnames(Safariland))
higher_attributes(Safariland.web)
# higher_attributes(Safariland.web)$rand
```

# Subsetting
```{r}
Safariland.web[1:5, 1:3]
Safariland.web[1:5, 1]
Safariland.web[1:5, 1] <- 1:5
Safariland.web[1:5, 1, drop = F]
Safariland.web[1:5, 1, drop = T]
Safariland.web[5, 3, drop = T]
Safariland.web[5, 3, drop = F]
```

# Test space
TODO: check whether names are taken with $ syntax.
```{r}
test.web <- bipartite_web(genweb(3, 5))
rand_v <- runif(5)
names(rand_v) <- c("col3", "col2", "col4", "col1", "col5")
print(rand_v)
higher_attributes(test.web)$rand <- rand_v
higher_attributes(test.web)
higher_attributes(test.web)$rand2 <- rand_v[colnames(test.web)]
higher_attributes(test.web)
```

```{r}
higher_attributes(Safariland.web)$rand2 <- runif(ncol(Safariland.web))
higher_attributes(Safariland.web)
```

```{r, max.print = 100}
options(max.print = 9999)
print(higher_attributes(Safariland.web))
```

```{r}
plot(Safariland.web, higher_color_attr = "rand2", higher_border = "black", link_border = "black")
```

```{r}
plot(Safariland.web,
     higher_color = c("Yellow", "Blue", "Green", "Red"),
     higher_color_attr = "Order",
     horizontal = TRUE)
```

```{r}
plot(Safariland.web,
     higher_color_attr = "Order",
     horizontal = TRUE)
```

```{r}
plot(Safariland.web,
     higher_color = c("Hymenoptera" = "Yellow",
                      "Diptera" = "Blue",
                      "Coleoptera" = "Green",
                      "Araneae" = "Red"),
     higher_color_attr = "Order",
     lower_color_attr = "Life.form",
     lower_color = rainbow(10),
     horizontal = TRUE)
```

## legend
```{r}
plot(Safariland.web,
     higher_color = c("Hymenoptera" = "Yellow",
                      "Diptera" = "Blue",
                      "Coleoptera" = "Green",
                      "Araneae" = "Red"),
     higher_color_attr = "Order",
     lower_color_attr = "Life.form",
     lower_color = rainbow(10),
     horizontal = TRUE,
     higher_border = "black",
     lower_border = "black",
     link_border = "black",
     mar = c(1,1,1,6))
par(xpd = T)
legend(1.2, 1,
       legend = c("Hymenoptera",
                  "Diptera",
                  "Coleoptera",
                  "Araneae"),
       fill = c("Yellow",
                "Blue",
                "Green",
                "Red"),
       title = "Order")
```

```{r}
plot(Safariland.web,
     higher_color = c("Hymenoptera" = "Yellow",
                      "Diptera" = "Blue",
                      "Coleoptera" = "Green",
                      "Araneae" = "Red"),
     higher_color_attr = "Order",
     lower_color_attr = "Life.form",
     lower_color = rainbow(10),
     horizontal = FALSE,
     higher_border = "black",
     lower_border = "black",
     link_border = "black",
     mar = c(1,1,1,6))
par(xpd = T)
legend(1, 1,
       legend = c("Hymenoptera",
                  "Diptera",
                  "Coleoptera",
                  "Araneae"),
       fill = c("Yellow",
                "Blue",
                "Green",
                "Red"),
       title = "Order")
```

## boolean attribute plot

```{r}
higher_attributes(Safariland.web)$hymenoptera <- higher_attributes(Safariland.web)$Order == "Hymenoptera"
plot(Safariland.web,
     higher_color = c("Black", "#ffbb00"),
     higher_color_attr = "hymenoptera",
     lower_color_attr = "Life.form",
     lower_color = hcl.colors(4, palette = "Dark 3"),
     horizontal = TRUE,
     sorting = "ca")
```

```{r}
plot(order_web(Safariland.web, "Order"),
     higher_color = rainbow(4),
     higher_color_attr = "Order",
     horizontal = TRUE)
```

```{r}
Safariland.web[, higher_attributes(Safariland.web)$Order == "Diptera"]

plot(Safariland.web[, higher_attributes(Safariland.web)$Order == "Diptera"],
     horizontal = TRUE,
     higher_color_attr = "Family",
     higher_color = rainbow(5),
     sorting = "ca")
```

# Bipartite webarrays

```{r}
warray <- bipartite_webarray(Safariland, vazarr)
print(warray)
```

```{r}
higher_attributes(warray, "Safariland")$rand <- runif(41)
print(warray)
higher_attributes(warray)$No <- 1:41
warray[,,apply(warray,3,sum) > 50]
warray[,higher_attributes(warray)$No > 10,]
warray[1:2,higher_attributes(warray)$No > 10,]
```

```{r}
warray[,,1]
```
